Potential and limits of site-specific substitution model inference.

### 2019-12-29 manuscript/ms.tex
edits to manuscript, added additional references

While biologically plausible, estimating such models from data exacerbates the over-fitting problem and it rarely attempted in practice.
In the context of the site-specific models this issue has become known as {\em extensive parametrization} or even {\em infinitely many parameters problem} \cite{Rodrigue557,spielman2016extensively}.
With sufficient data, however, site specific parameters can be accurately estimated \citep{tamuri_estimating_2012,spielman2016extensively,scheffler2014validity}.
Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on model deviations.

\begin{figure*}[htb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha1.5}.pdf}
	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
	The estimates are vastly more accurate than naive estimates from the state frequencies in alignment columns. Different curves are for $n\in [100,300,1000,3000]$ sequences per tree.
	% (B) While the $\eqp_i^a$ are accurately inferred by the linearized additive procedure with known internal states, the mutation rates are systematically underestimated, as expected.
	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates. Regularization via pseudo-counts reduces over-fitting at low divergence.  }
	\label{fig:dressed}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Model
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Efficient inference of site specific substitution models}
Following previous authors \citet{halpern_evolutionary_1998}, we assume that the substitution matrix $W_{ij}$ between states $j,i$ does not depend on the position in the sequence, while
the overall rates and the equilibrium frequencies depend on the site.
Such a site-specific general time-reversible (GTR)) model, can be parameterized as:
\begin{eqnarray}
Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
\label{eq:Qij}
\end{eqnarray}
where $W_{ij}$ is a symmetric matrix with $W_{ii}=0$ and the second equation ensures conservation of probability.
In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to ensure that the average rate per site is $\mu^{a}$.

Extending the iterative expectation maximization approach by \citet{bruno1996modeling}, we show in the appendix that the iterative update rules
\begin{equation}
\label{eq:update}
	\begin{split}
		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
	\end{split}
\end{equation}
approximately maximize the likelihood of the data.
Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree and $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$
$\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ divided by the total tree length.
To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated. This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ACCURACY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Accuracy of inferences}
The validity of this iterative solution will depend on the accuracy of the linear approximation made, the degree to which $n_{ij}^b$ and $\tau_{j}^b$ can be estimated, and the accuracy of the tree reconstruction.
In addition, the most likely model parameters will deviate from the true parameters due to over-fitting.

To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites.
We quantify the accuracy of the inference as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities and compare inference using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_nuc_ratealpha1.5}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_aa_ratealpha1.5}.pdf}
	\caption{{\bf Quantification of errors stemming from tree inference and ancestral reconstruction.}
	The left and right panels show the mean-squared deviation $\chi^2$ of inferred $\hat{\eqp}_i^a$ from the true $\eqp_i^a$ for 4-letter and 20-letter alphabets, respectively.
	At large root-to-tip distances, ancestral reconstruction becomes less and less certain and estimation of $\eqp_i^a$ fails (red lines).
	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch length using the updated models (or using the true tree, yellow/pink).
	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=3$.}
	\label{fig:reconstructed}
\end{figure*}

Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site for estimates of $\eqp_i^a$ from the alignment and using the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
The plot is shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope 1 implies that the accuracy is inversely proportional to the expected number of substitutions.
This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
This scaling becomes more accurate with less rate variation.
At the largest substitution rates, branch length are on the order of 0.5 and the linear approximation underlying the iterative equations is not longer accurate.
Nevertheless, the accuracy of the $\hat{\eqp}_i^a$ continues to increase.
By contrast, using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry.

The accuracy of substitution rates is affected by linearization and Eq.~\ref{eq:update} will consistently underestimate $\mu^b$.
This is expected as the linearization ignores cases where the same site changes twice along a branch in very much the same way as Hamming distance will underestimate branch length.
The relative substitution rates, however, are accurately estimated (with Pearson correlation coefficients $>0.9$ as soon as the majority of sites experience several mutations across the tree), see Fig.~\ref{fig:dressed}B.

@@

\subsection*{Ignoring site specific frequencies results in dramatic branch length underestimates}
As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^a}^2$ is increasing sharply with more peaked preferences.
Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent rate \citet{wertheim_purifying_2011}.

@@

%% dependence on pseudo count
Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted tree (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
While branch length are accurately estimated when using the true model, they are systematically underestimated with increasing pseudo-counts and or without site specificity.
Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogeny which are particularly prone to underestimation.
When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
This effect is entirely due to skewed equilibrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^a=q^{-1}$ with the same rate variation.
%% dependence on nucleotide and amino acid skew
The underestimation of branch lengths is less of a problem when using larger alphabets and the true probability are not too heavily skewed, see Fig.~\ref{fig:tree_length_supp}.

\begin{figure}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation}.pdf}
	\caption{{\bf Moderate model inaccuracies result in large substantial branch length underestimation}.
	}
	\label{fig:model_deviation}
\end{figure}

Surprisingly, the inferred site specific model only partially rectifies the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
The principle contributor to this deviation are inaccuracies in the rate estimates and combining the true rates with inferred preferences largely eliminates the error in branch length estimation (see Fig.~\ref{fig:tree_length}, lines labeled ``(true rates)''.
%If, however, preferences are estimated poorly for example through excessive regularization ($\pc=10$ in Fig.~\ref{fig:tree_length}), using the correct rates hardly improves the estimate.
%The latter is consistent with the observation that standard ML tree inference using IQ-tree with a GTR+R10 model severely underestimates branch lengths.

\subsection*{Minor model deviations can have substantial effects on divergence estimates}
In the previous section, we have observed that even small model deviations, for example those introduced by strong regularization, already result in substantial mis-estimation.
To more systematically investigate this effect, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$ as follows.
For a mixing fraction $\alpha$, we constructed a model with rates
\begin{equation}
	\mu^a = \alpha \langle \mu^a\rangle + (1-\alpha) \mu^a_{\mathrm{true}}
\end{equation}
where $\langle \mu^a\rangle$ is the expectation value of the rate.
Mixtures of site-specific frequencies are constructed analogously.

Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
Deviation in rate estimates result immediately in underestimated branch length, while substantial effect of too flat site specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
If the same preferences are used for every site ($\alpha=1$), however, deviations are substantial.
Deviations in rate and preferences are approximately additive.
These observations are consistent with the finding above using inferred preferences and true substitution rates result in much more accurate branch length estimates than inferring both rates and preferences.

@@

\subsection*{Applications to large HIV alignments}
Since its zoonosis, HIV-1 has diversified into several different subtypes that differ from each other at about 20\% of sites in their genome.
For each subtype, thousands of sequences are available, which typically differ by about 10\% \citep{LANL}.
This large sample of moderately diverged sequences is a unique opportunity to apply our iterative inference frame work.
We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site specific GTR models at the nucleotide and amino acid level.

One of the original motivations for site-specific models was the idea that equilibrium frequencies reflect the relative fitness of the different states at each position \citep{halpern_evolutionary_1998}.
Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ on the fitness different $s^a_{ij}$ and the effective population size $N$ as
\begin{equation}
	f^a_{ij} = \frac{1-e^{-2s^a_{ij}}}{1-e^{-2Ns^a_{ij}}} \approx
	\begin{cases}
		2s^a_{ij} & s^a_{ij}>0 \\
		2s^a_{ij} e^{2Ns^a_{ij}} & s^a_{ij} < 0
	\end{cases}
\end{equation}
On longer time scales, the fixation rates $f^a_{ij}$ should correspond to transitions rates of the site specific GTR model $\mu^a Q^a_{ij}$.
The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations.
Nevertheless, the logarithm of the ratio $\log f^a_{ij}/f^a_{ji} \approx 2Ns^a_{ij}$ is an interpretable quantity related to the average fitness difference between states on time scales longer than the population genetic scale $N\sim T_c$.
We generalize this notion to multiple states and define a fitness score of state $i$ at position $a$ as the ratio of rates into and out of state $i$
\begin{equation}
	e^{2N s_i^a} = \frac{\Gamma_{in}}{\Gamma_{out}} = \frac{\sum_j \pi_i^a W_{ij}}{\sum_j W_{ij}\pi_j^a}
\end{equation}
The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the multiplicative factor $2N$ is unknown.
We compare this proxy of fitness differences to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
However, instead of a linear relation ship between $s_i^a$ and within host estimates of fitness costs, we observe a linear relationship between the logarithm of fitness effects measured with-in host and $s_i^a$ (
see Fig.~\ref{fig:HIV_fitness})
This relationship explains about half the variance of nucleotide effects and slightly less of amino acid effects.

The fact that this relationship deviates from the expectation
\begin{equation}
	\nonumber
	\textrm{cross-sectional} \sim \textrm{with-in host}
\end{equation}
and instead is approximately
\begin{equation}
	\nonumber
	\textrm{cross-sectional} \sim \log\left(\textrm{with-in host}\right)
\end{equation}
points towards different process that drive cross-sectional and with-in host diversification.
While the order of fitness effects with-in host and cross-sectionally seem to be mostly concordant, with-in host variation is much greater than cross-sectionally.
Within host fitness effects are masked and damped at the population level, possibly due to fluctuating by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
The ratio of in/out rates $e^{-2Ns_i^a}$ is, however, a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_fitness}).

@@

Next, we quantified the effect that ignoring site-specific preferences has on branch length estimates.
We generated sequence pairs that evolved for a specific time $t$ under the inferred site-specific model and then inferred a maximum likelihood branch length between these two sequences using a model with the average preferences and the same per site rate variation.
As soon as the length of the simulated branch approaches 0.2, the length inferred by a model without site-specific preferences deviates substantially from the true value and saturates around 0.5 for larger and larger $t$.
As expected, these deviations are even more severe when estimating branch length as simple sequence divergence (p-distance), while using the generating model reproduces the correct branch length.

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.8\textwidth]{{../figures/HIV_B_pol_branch_length_error_pc_0.010}.pdf}
	\caption{{\bf Underestimation of divergence in HIV.}
	  Panel A shows the estimated ML branch length for the model used to generate the sequences and a model with constant equilibrium frequencies as a function of true branch length. For comparison, the p-distance between the two simulated sequences is also shown. The top axis shows branch length in units of years assuming a substitution rate of 0.002/year. Error bars denote one standard deviation.
      Panel B shows the distribution of rates across sites for both models. They differ slightly in the overall scale since rates have been rescaled such that the average substitution rate in both models is identical.}
	\label{fig:HIV_branch_length}
\end{figure*}

@@

While accurate estimates of the site-specific GTR models are a possible given a large enough set of mildly diverged sequences, joint inference of model and branch lengths remains difficult.
This difficulty can be largely traced down to poorly constrained estimates of site-specific rates -- one primary reason why these rates are typically not estimated but treated as random effects \citep{yang1996among}.

>>>>

While biologically plausible, estimating such models from data exacerbates the over-fitting problem and it rarely attempted in practice.
In the context of the site-specific models this issue has become known as {\em extensive parametrization} or even {\em infinitely many parameters problem} \cite{Rodrigue557,spielman2016extensively}.
With sufficient data, however, site specific parameters can be accurately estimated \citep{tamuri_estimating_2012,spielman2016extensively,scheffler2014validity}.
Here, we implement an EM-style algorithm inspired by \citep{bruno1996modeling} to infer site-specific rates and preferences from simulated data, quantify its accuracy and the different sources of bias and noise, and show how divergence time estimates depend on the fidelity with which site specific model parameters are known.

\begin{figure*}[htb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha1.5}.pdf}
	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\eqp_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
	The estimates are more accurate than naive estimates from the state frequencies in alignment columns. The advantage over naive estimates increases with data set size as seem from different curves with $n\in [100,300,1000,3000]$ sequences per tree.
	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates. Regularization via pseudo-counts reduces over-fitting at low divergence. Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$. }
	\label{fig:dressed}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Model
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Efficient inference of site specific substitution models}
Following previous \citet{halpern_evolutionary_1998}, we assume that the substitution matrix $W_{ij}$ between states $j,i$ does not depend on the position in the sequence, while
the overall rates and the equilibrium frequencies depend on the site.
Such a site-specific general time-reversible (GTR)) model, can be parameterized as:
\begin{eqnarray}
Q^{a}_{ij} &=& \mu^{a}\eqp^{a}_{i} W_{ij} \textrm{ for } i\neq j,\nonumber \\
Q^{a}_{ii} &=& -\sum_k Q^{a}_{ki}
\label{eq:Qij}
\end{eqnarray}
where $W_{ij}$ is a symmetric matrix with $W_{ii}=0$ and the second equation ensures conservation of probability.
In addition, we require $\sum_i \eqp^{a}_i = 1$ and $\sum_{a=1}^L\sum_{i\neq j}W_{ij}p^{a}_ip^{a}_j=L$ to ensure that the average rate per site is $\mu^{a}$.

Extending the iterative expectation maximization approach by \citet{bruno1996modeling}, we show in the appendix that the iterative update rules
\begin{equation}
\label{eq:update}
	\begin{split}
		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
	\end{split}
\end{equation}
approximately maximize the likelihood of the data.
Here $\tau^b_j$ is the time site $b$ spends in state $j$ across the tree and $n^b_{ij}$ are the number of transitions from state $j$ to state $i$ at site $b$
$\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ divided by the total tree length.
To make the behavior of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated. This (i) shows that the update rules are multiplicative and therefore ensure positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
These update rules are an example of non-negative factorization algorithms \citep{lee_algorithms_2001}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ACCURACY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Accuracy of inferences}
The validity of this iterative solution will depend (i) on the accuracy of the linear approximation, (ii) the accuracy to which $n_{ij}^b$ and $\tau_{j}^b$ can be estimated, and (iii) the accuracy of the tree reconstruction.
In addition, the Maximum Likelihood estimates might deviate from the true parameters due to over-fitting.

To assess these sources of error independently, we simulated sequences evolving along a fixed tree and explicitly recorded $n_{ij}^b$ and the $\tau_j^b$ for a range of sample sizes, levels of divergence, and different degrees of variation between sites, see Materials and Methods.
We quantify the accuracy of the inference as $\chi^2 = L^{-1}\sum_{a,i}(\hat{\eqp}_i^a - \eqp_i^a)^2$ where $\hat{\eqp}_i^a$ and $\eqp_i^a$ are the inferred and true equilibrium probabilities and compare inference using (i) the true ancestral sequences, (ii) reconstructed ancestral sequences, and (iii) reconstructions on the true phylogenetic tree.

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_nuc_ratealpha1.5}.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_rtt_aa_ratealpha1.5}.pdf}
	\caption{{\bf Quantification of errors stemming from tree inference and ancestral reconstruction.}
	Panels A and B show the mean-squared deviation $\chi^2$ of inferred $\hat{\eqp}_i^a$ from the true $\eqp_i^a$ for 4-letter and 20-letter alphabets, respectively.
	At large root-to-tip distances, ancestral reconstruction becomes less and less certain and estimation of $\eqp_i^a$ fails (red lines).
	These errors are gradually eliminated by first summing over ancestral uncertainty (violet), iteratively redoing ancestral reconstruction using the inferred model (brown), and re-optimizing branch length using the updated models (or using the true tree, yellow/pink).
	Data in this figure uses was generated assuming Gamma distributed rate variation with $\alpha=1.5$.}
	\label{fig:reconstructed}
\end{figure*}

Fig.~\ref{fig:dressed} shows the average squared error $\chi^2$ for a trees with $n\in [100,300,1000,3000]$ taxa and a range of substitution rates as a function of the expected number of substitutions per site for estimates of $\eqp_i^a$ from the alignment and using the iterative scheme with correct $n_{ij}^a$ and $\tau_i^a$ as input.
The plot is shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope 1 implies that the accuracy is inversely proportional to the expected number of substitutions.
This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.
Deviations of the scaling $\chi_2\sim 1/\textrm{tree length}$ are due to rate variation.
The data shown in Fig.~\ref{fig:reconstructed}, assumes that rates are Gamma distributed with $\alpha=1.5$ and variation in inference accuracy due to rate variation increases the average error.
Increasing $\alpha$ to 3 improves this scaling.
A naive approach using the frequencies of different sequences states in the alignment as estimates for $\eqp_i^b$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:reconstructed}A).

At the largest substitution rates, branch lengths are on the order of 0.5 and the linear approximation underlying the iterative equations is not longer accurate.
Nevertheless, the accuracy of the $\hat{\eqp}_i^a$ continues to increase.
While equilibrium frequencies are insensitive, the substitution rate estimates are affected by linearization and Eq.~\ref{eq:update} will consistently underestimate $\mu^b$.
This is expected as the linearization ignores cases where the same site changes twice along a branch in very much the same way as Hamming distance will underestimate branch length.
The relative substitution rates, however, are accurately estimated (with Pearson correlation coefficients $>0.9$ as soon as the majority of sites experience several mutations across the tree), see Fig.~\ref{fig:dressed}B.

@@

\subsection*{Ignoring site specific frequencies results in branch length underestimates}
As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{\eqp_i^a}^2$ is increasing sharply with more peaked preferences.
Models that don't account for site specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent rate \citet{wertheim_purifying_2011}.

@@

%% dependence on pseudo count
Fig.~\ref{fig:tree_length} shows the average branch length (panel A) and the average root-to-tip distance of mid-point rooted trees (panel B) with branch length optimized using (i) the true model, (ii) using the GTR+R10 model of IQ-tree, and (iii) inferred models using different degrees of regularization.
While branch lengths are accurately estimated when using the true model, they are systematically underestimated by the GTR=R10 model without site specificity.
Underestimation is particularly severe for the average root-to-tip distance which is dominated by long branches deep in the phylogeny which are prone to underestimation.
When ignoring site-specific preferences, the inferred root-to-tip distance is essentially independent from the true value for distances greater than 1 (Fig.~\ref{fig:tree_length}B).
This effect is entirely due to skewed equilibrium frequencies, as branch length inference by IQ-tree is accurate if the simulated data had flat $\eqp_i^a=q^{-1}$ with the same rate variation.
%% dependence on nucleotide and amino acid skew
The underestimation of branch lengths is less of a problem when using larger alphabets and the true probability are not too heavily skewed, see Fig.~\ref{fig:tree_length_supp}.

\begin{figure}[tb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/model_deviation_n300}.pdf}
	\caption{{\bf Sensitivity of branch length estimates on model misspecification}.
	Panels A \& B show the relative error in total tree length when using a mixture model as defined in Eq.~\ref{eq:mixture_model} for branch length inference. Panel A shows this error as a function of the mixing fraction $\alpha$ for $\langle \mu \rangle = 0.2$. Panel B shows the error as a function of the evolutionary rate $\langle \mu \rangle$ for $\alpha=1$.
	The mixing is applied to the equilibrium frequencies $\eqp_i^a$, the rates $\mu^a$, or both.
	}
	\label{fig:model_deviation}
\end{figure}

Surprisingly, the inferred site specific model only partially rectifies the problem of branch length under-estimation, despite the fact that these models are close to the true model in terms of low $\chi^2$.
The principle contributor to this deviation are inaccuracies in the rate estimates.
Combining the true rates with inferred preferences reduces the error in branch length estimation (see Fig.~\ref{fig:tree_length}, lines labeled ``(true rates)''.

\subsection*{Minor model deviations can have substantial effects on divergence estimates}
In the previous section, we have observed that model deviations, for example the error made during model inference, can result in substantial errors in branch length estimates.
To investigate this effect more systematically, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$ as follows:
For a mixing fraction $\alpha$, we constructed a model with rates
\begin{equation}
\label{eq:mixture_model}
	\mu^a = \alpha \langle \mu^a\rangle + (1-\alpha) \mu^a_{\mathrm{true}}
\end{equation}
where $\langle \mu^a\rangle$ is the expectation value of the rate.
Mixtures of site-specific frequencies are constructed analogously.

Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
Deviation in rate estimates result immediately in underestimated branch length, while substantial effect of too flat site specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
If the same preferences are used for every site ($\alpha=1$), however, deviations are substantial.
Deviations in rate and preferences are approximately additive.
These observations are consistent with the finding above using inferred preferences and true substitution rates result in more accurate branch length estimates than inferring both rates and preferences.

@@

\subsection*{Applications to large HIV alignments}
Since its zoonosis, HIV-1 has diversified into several different subtypes that differ from each other at about 20\% of sites in their genome.
For each subtype, thousands of sequences are available, which typically differ by about 10\% \citep{LANL}.
This large sample of moderately diverged sequences should be suited to apply our iterative inference frame work.
We downloaded alignments of HIV-1 pol sequences from the LANL data base, constructed phylogenetic trees, and inferred site specific GTR models at the nucleotide and amino acid level, see Materials and Methods.

One of the original motivations for site-specific models was the idea that equilibrium frequencies reflect the relative fitness of the different states at each position \citep{halpern_evolutionary_1998}.
Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ on the fitness different $s^a_{ij}$ and the effective population size $N$ as \citep{kimura_diffusion_1964}
\begin{equation}
	f^a_{ij} = \frac{1-e^{-2s^a_{ij}}}{1-e^{-2Ns^a_{ij}}} \approx
	\begin{cases}
		2s^a_{ij} & s^a_{ij}>0 \\
		2s^a_{ij} e^{2Ns^a_{ij}} & s^a_{ij} < 0
	\end{cases}
\end{equation}
On longer time scales, the fixation rates $f^a_{ij}$ should correspond to transitions rates of the site specific GTR model $\mu^a Q^a_{ij}$.
The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations.
Nevertheless, the logarithm of the ratio $\log f^a_{ij}/f^a_{ji} \approx 2Ns^a_{ij}$ is an interpretable quantity related to the average fitness difference between states on time scales longer than the population genetic scale $N\sim T_c$.
We generalize this notion to multiple states and define a fitness score of state $i$ at position $a$ as the ratio of rates into and out of state $i$
\begin{equation}
	e^{2N s_i^a} = \frac{\Gamma_{in}}{\Gamma_{out}} = \frac{\sum_j \pi_i^a W_{ij}}{\sum_j W_{ij}\pi_j^a}
\end{equation}
The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the multiplicative factor $2N$ is unknown.
We compared this proxy of fitness differences to fitness costs measured using mutation selection balance models and with-in host diversity data of HIV \citep{zanini2017vivo}.
However, instead of a linear relation ship between $s_i^a$ and within host estimates of fitness costs, we observed a linear relationship between the logarithm of fitness effects measured with-in host and $s_i^a$ (
see Fig.~\ref{fig:HIV_fitness})
This relationship explains about half the variance of nucleotide effects and about one third of amino acid effects.

The fact that this relationship deviates from the expectation
\begin{equation}
	\nonumber
	\textrm{cross-sectional} \sim \textrm{with-in host}
\end{equation}
and instead is approximately
\begin{equation}
	\nonumber
	\textrm{cross-sectional} \sim \log\left(\textrm{with-in host}\right)
\end{equation}
points towards different process that drive cross-sectional and with-in host diversification.
While the order of fitness effects with-in host and cross-sectionally seem to be mostly concordant, with-in host variation is much greater than cross-sectional variation.
Within host fitness effects are masked and damped at the population level, possibly due to fluctuating by diverse host immune systems or epistasis \citep{zanini2015population,shekhar_spin_2013}.
The ratio of in/out rates $e^{-2Ns_i^a}$ is, however, a very good correlate of alignment diversity (see Fig.~\ref{fig:HIV_diversity_supp}).

@@

Next, we quantified the effect of ignoring site-specific preferences on branch length estimates.
We generated sequence pairs that evolved for a specific time $t$ under the inferred site-specific model and then inferred a maximum likelihood branch length between these two sequences using a model with homogeneous equilibrium frequencies set to the average frequencies across sites while maintaining site-specific rate variation.
As soon as the length of the simulated branch approaches 0.2, the length inferred by a model without site-specific preferences deviates substantially from the true value and saturates around 0.5 for larger and larger $t$.
As expected, these deviations are even more severe when estimating branch length as simple sequence divergence (p-distance), while using the generating model reproduces the correct branch length.

Assuming a typical evolutionary rate of HIV of 0.002 changes per site and year, this analysis suggests that length estimates of branches longer than 100 years start to become inaccurate.
Furthermore, this analysis suggests very little signal to estimate the length of branches that are longer than 300y.

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.8\textwidth]{{../figures/HIV_B_pol_branch_length_error_pc_0.010}.pdf}
	\caption{{\bf Underestimation of divergence in HIV.}
	  Panel A shows the estimated ML branch length for the model used to generate the sequences and a model with constant equilibrium frequencies as a function of true branch length. For comparison, the p-distance between the two simulated sequences is also shown. The top axis shows branch length in units of years assuming a substitution rate of 0.002/year and site. Error bars denote one standard deviation.
      Panel B shows the distribution of rates across sites for both models.
      About 20\% of sites are essentially invariable, while the rates of the remainder vary by at least 10fold.
      The distributions differ slightly in the overall scale since rates have been rescaled such that the average substitution rate in both models is identical.}
	\label{fig:HIV_branch_length}
\end{figure*}

@@

While accurate estimates of the site-specific GTR models are a possible given a large enough set of mildly diverged sequences, joint inference of model and branch lengths remains difficult.
This difficulty can be largely traced down to poorly constrained estimates of site-specific rates -- one  reason why these rates are typically not estimated but treated as random effects \citep{yang1996among}.
